# Data Visualization with `ggplot`

> A great resource for data visualuation in `R` is the [**R Graph Gallery**](http://r-graph-gallery.com/). The examples and information below is a small sample of `R` data visualization basics.

## `ggplot`

`ggplot2` (referred to as `ggplot`) is a powerful graphics package that can be used to make very impressive data visualizations (see [contributions to #TidyTueday on Twitter](https://nsgrantham.shinyapps.io/tidytuesdayrocks/), for example). The following examples will make use of the Learning R Survey data, which has been partially processed (Chapters 2 and 3) and the `palmerpenguins` data set, as well as several of datasets included with `R` to show the basic principles of using `ggplot`. Then, we will put these basics together to make several beautiful visualizations.

```{r package_setup_viz, message=FALSE, warning=FALSE}
library(tidyverse)
# devtools::install_github("allisonhorst/palmerpenguins")
library(palmerpenguins)
data(penguins)
load("data/rsurvey.Rdata")
```

## Grammar of Graphics

The "gg" of `ggplot` refers to the "grammar of graphics". For `ggplot`, this means a visualization must have specific elements to make a complete graphic, just as an utterance or written line must have specific elements to make a grammatically correct sentence.

A simple plot contains the following elements:
```{r}
ggplot(data=rsurvey)+       # reference to data
  geom_bar(                 # geometry - a shape
    aes(x= qr_experience,   # aesthetics - x, y, and color values
        fill=qr_experience)   
  )
```


## Data

There are several ways to refer to a data object in `ggplot`. You can call data within `ggplot` (e.g. `ggplot(rsurvey)` ) or you can call it outside of `ggplot` within a `dplyr` chain (e.g `rsurvey %>% ggplot()`). The advantage of this is you can easily manipulate data directly into `ggplot` without saving it as a data object:

```{r}
rsurvey %>% # data 
  filter(qr_experience == "Intermediate") %>% #manipulation
  ggplot()+ #visualization
  coord_flip()+   # this makes the plot horizontal
  geom_bar( 
    aes(x=qr_how_often_used)
  )
```

You can also call data for individual shapes. This would allow you to use different data objects to form your graphic, or to use the same data object but filter it to display different information. The following graphic demonstrates this. 

*Please note this graphic is for demonstration purposes only and does not represent a useful data visualization.*

```{r message=FALSE, warning=FALSE}
ggplot()+
  # data for beginners only - bar chart
  geom_bar(data=rsurvey %>%
             filter(qr_experience == "Beginner"), 
             aes(x=age,
                 fill = qr_experience))+
  # data for experts - line chart
  geom_line(data=rsurvey %>%
             filter(qr_experience == "Expert"),
             aes(x=age,
                 color = qr_experience), stat="count")
```

> A note about `+`
>
> Across the `Tidyverse`, the `%>%` "pipe operator" is used to chain commands into simple codes. It stands for "and then". However, in `ggplot`, despite being part of the tidyverse, different elements are connected with the plus sign `+`.

## Aesthetics

Aesthetics are the way you connect data to the elements inside the graphic. Aesthetics tell `ggplot` what should be on the x-axis, what should be on the y-axis, and what the colors should be.

Different geometries (shapes) may have different aesthetics, but x, y, and color/fill are the most common.

  * `color=` is used for:
    + `geom_point()` - dots, circles, scatterplots
    + `geom_line()` - line charts
  * `fill=` is used for:
    + `geom_col()` / `geom_bar()` - column/bar charts
    + `geom_area()` - area charts
    
Using `color=` or `fill=` to refer to a categorical variable (called a  "discrete" variable in `ggplot`) allows you to separate the shape by that category. Here is an example with and without specifiying a color:

```{r}
library(patchwork) # for combining two plots
(ggplot(penguins)+
  geom_bar(aes(x=species))+
    ggtitle(label = "No `fill =`"))+
(ggplot(penguins)+
  geom_bar(aes(x=species, fill=sex))+
       ggtitle(label = "fill = sex"))
```


## Geometries

Geometries are the different shapes one can make using `ggplot`. They all start with `geom_` and can be stacked together by simply using `+`. The first `geom` is always first layer and any additional layers are stacked on top of it. (See [Lollipop Charts][Lollipop Charts] for an example.)


### Bar Charts

You can make bar charts with either `geom_bar()` or `geom_col()`.

#### `geom_bar`

`geom_bar` requires:

  * an x-value and is useful if you are just getting **a count of the data*
  
```{r}
ggplot(penguins)+
  geom_bar(aes(x=species))
```

`geom_bar` may also have:

  * a y-value and a `stat=` value if you want to specify how the y-value data should be shown
    + `stat="identity"`- gives a sum of all the values of y
    + `stat="summary"` - gives a mean of the values of y

Compare: 

```{r fig.width=7.5, message=FALSE, warning=FALSE}
(ggplot(penguins)+
  geom_bar(aes(x=species, y=bill_length_mm),
            stat = "identity")+
  labs(title='stat="identity" produces a sum'))+
(ggplot(penguins)+
  geom_bar(aes(x=species, y=bill_length_mm),
            stat = "summary")+
  labs(title='stat="summary" produces a mean'))
```

#### `geom_col`

`geom_col` requires x and y values. It *does not* use `stat=`.

```{r message=FALSE, warning=FALSE}
penguins %>% # data
  group_by(species) %>% # manipulation
  summarize(avg = mean(bill_length_mm, na.rm=T)) %>% #manipulation
  ggplot()+ 
  geom_col(aes(x=species, y=avg))
```



#### Horizontal Bar Chart

To make a horizontal bar chart, add `coord_flip()`:

```{r}
ggplot(penguins)+ # data
  geom_bar(aes(x=species, y=bill_length_mm), #bar chart
            stat = "summary")+
  coord_flip() # make it horizontal
```

*note:* `coord_flip()` can be placed anywhere after `ggplot()`

#### Stacked Bar Chart

To make a stacked bar chart, include `fill=`

```{r}
penguins %>% # data
  filter(!is.na(sex)) %>% # manipulation
ggplot()+ # ggplot
  geom_bar(aes(x=species, y=bill_length_mm,  # bar chart
               fill=sex), # separate by sex
            stat = "summary")
```

#### 100% Stacked Bar Chart

To make a **100% stacked** bar chart, include `fill=` and `position="fill"` after `aes()`

```{r message=FALSE, warning=FALSE}
penguins %>% #data
  filter(!is.na(sex)) %>% #manipulation
ggplot()+ 
  geom_bar(aes(x=species, y=bill_length_mm, #bar chart
               fill=sex), # separate by sex
            stat = "summary",
           position="fill") # make it 100%
```

#### Side-by-Side Bar Chart

To make a side-by-side bar chart, include `fill=` and `position="dodge"` after `aes()`

```{r message=FALSE, warning=FALSE}
penguins %>% # data
  filter(!is.na(sex)) %>% # manipulation
ggplot()+
  geom_bar(aes(x=species, y=bill_length_mm, 
               fill=sex), # separate by category
            stat = "summary", 
           position="dodge") # make it side-by-side
```

### Histograms

Histograms can be made with `geom_histogram`. They only require an x-value. You can decide the bin width by adding `binwidth=` after `aes()`

```{r message=FALSE, warning=FALSE}
ggplot(penguins)+
  geom_histogram(aes(x=flipper_length_mm))+
  labs(title="No bin width specified")

ggplot(penguins)+
  geom_histogram(aes(x=flipper_length_mm),
                 binwidth=10)+
  labs(title="Bin width = 10")
  
ggplot(penguins)+
  geom_histogram(aes(x=flipper_length_mm),
                 binwidth=1)+
  labs(title="Bin width = 1")
```


### Boxplots

A boxplot uses `geom_boxplot()`. It requires x-values. Y-values are optional but useful if you want to compare multiple boxplots.

*Note:* Use `coord_flip()` to make it easier to read.

```{r message=FALSE, warning=FALSE}
ggplot(penguins)+
  coord_flip()+
  geom_boxplot(aes(x=flipper_length_mm, y=sex))
```


### Scatterplots

We can use `geom_point()` for scatterplots. This requires x and y values, both continuous.

```{r message=FALSE, warning=FALSE}
ggplot(penguins)+
  geom_point(aes(x=bill_length_mm, y=flipper_length_mm))
```

#### Point Size

Point size can be based on a single number or the data itself.

You can control **overall** point size by adding `size=` **after** `aes()`:

```{r message=FALSE, warning=FALSE}
ggplot(penguins)+
  geom_point(aes(x=bill_length_mm, y=flipper_length_mm),
             size=3)
```


You can make the values in the data also determine point size by using `size=` **inside** `aes()`:

```{r message=FALSE, warning=FALSE}
ggplot(penguins)+
  geom_point(aes(x=bill_length_mm, y=flipper_length_mm,
                 size=flipper_length_mm))
```

#### Point Color

You can differentiate the dots by adding `color=`

```{r message=FALSE, warning=FALSE}
ggplot(penguins)+
  geom_point(aes(x=bill_length_mm, y=flipper_length_mm,
                 color=species))+
  labs(title="Color added")

ggplot(penguins)+
  geom_point(aes(x=bill_length_mm, y=flipper_length_mm,
                 color=species,
                 size=flipper_length_mm), alpha=.6)+
  labs(title="Color and size based on data")
```

#### Shapes

Shapes are controlled like color and size. Inside `aes()` means that shapes are mapped to data. Outside `aes()` means there is one shape.

```{r message=FALSE, warning=FALSE}
ggplot(penguins)+
  geom_point(aes(x=bill_length_mm, 
                 y=flipper_length_mm,
                 shape=sex,
                 color=sex), size=5, alpha=.5)+
  labs(title="Shape and color controlled by data \nSize set outside aesthetics")

ggplot(penguins)+
  geom_point(aes(x=bill_length_mm, y=flipper_length_mm,
                 color=species),
             size=4, shape=3)+
  labs(title="Color controlled by data \nSize and shape set outside aesthetics")
```

> **A note on `alpha=`
>
>`alpha` is called after `aes()` to set the transparency of overlapping shapes. An alpha of `0` is complete transparency while an alpha of `1` is no transparency. An alpha of `.5`, set above, is 50% transparency.

#### Scatterplots for Categorical Variables

If you want to make a scatterplot for categorical variables, you will simply get a line of dots for each variable unless you use `geom_jitter()`, which adds random fluctuation in the variables.

Compare:

```{r fig.width=7.5, message=FALSE, warning=FALSE}
(ggplot(penguins)+
  geom_point(aes(x=bill_length_mm, y=species,
                 color=species))+
  labs(title="Using geom_point()")+
   theme(legend.position = "none"))+
  (ggplot(penguins)+
  geom_jitter(aes(x=bill_length_mm, y=species,
                 color=species))+
  labs(title="Using geom_jitter()")+
    theme(legend.position = "none"))
```

You can also use `position_jitter(width = NULL, height = NULL, seed = NA)` inside of `geom_point()` to achieve a similar effect.


### Barbell Charts

Barbell charts compare plot two related variables with a dot and show the distance between them with a line.

You can combine `geom_point()` with `geom_linerange()` to make a simple lollipop chart. `geom_linerange()` should be called first, as it must go below the dots layer for its line ends to be hidden by the dot. First, we will summarize the penguin data and then compare.

The following code builds the graphic by combining different data layers and different geometry layers.

```{r}
ggplot()+
  coord_flip()+
  geom_linerange(data=penguins %>%
                   group_by(species, sex) %>%
                   summarize(body_mass = mean(body_mass_g, na.rm=T)) %>%
                   drop_na(sex) %>%
                   pivot_wider(names_from = sex, values_from=body_mass),
    aes(x=species, ymin=male, ymax=female))+
  geom_point(data=penguins %>%
                  group_by(species, sex) %>%
                  summarize(body_mass = mean(body_mass_g, na.rm=T)) %>%
                  drop_na(sex),
             aes(x=species, y=body_mass, color=sex),
             size=5)
```

### Line Charts

You can use `geom_line()` for line charts to display values over time.
`geom_line()` requires an additional `group=` aesthetic. If there should be only 1 line because there is only 1 time variable, then use `group=1`. If you want to split the lines based on another variable, use `group=variable_name`.

For the below example, we will use the `AirPassengers` data that comes with `R` and transform it into a dataframe following an example from [StackOverflow](https://stackoverflow.com/questions/5331901/transforming-a-time-series-into-a-data-frame-and-back)

A line graph displaying a single line for year

```{r}
data(AirPassengers)

airpassengers <- data.frame(AirPassengers, year = trunc(time(AirPassengers)), 
month = month.abb[cycle(AirPassengers)])

airpassengers %>%
  group_by(year) %>%
  summarize(sum = sum(AirPassengers, na.rm=T)) %>%
  ggplot()+
  geom_line(aes(x=year, y=sum, group=1))

```

A line graph displaying 1 line per month

```{r}
ggplot(airpassengers)+
  geom_line(aes(x=year, y=AirPassengers, group=month))
```
We can add labels to the ends of the line using `geom_label()` (see [Labels][Labels]) but the lines are very close together, so we will use `ggrepel()` instead. This gives the labels space and connects them with their lines.

```{r}
library(ggrepel)

ggplot(airpassengers)+
  geom_line(aes(x=year, y=AirPassengers, group=month))+
  geom_label_repel(data=airpassengers %>% #use only the last year in the data set
                    filter(year == max(year)),
                  aes(x=year, y=AirPassengers, label=month))
```


### Colors

For colors related to values in a data set, see [Aesthetics][Aesthetics]

You can change the color of all the chart elements of a geometry using `fill=` outside of `aes()`. Here, color is not mapped to the data, thus it is inside `geom_col` but not in `aes()`. You can use `R` color names (e.g. "blue", "black", "grey80"), hex values (e.g. "#cccccc" or "#a85001", or RGB values (e.g. `rgb(0, 155, 255)`).

```{r}
penguins %>%
  group_by(species) %>%
  summarize(avg = mean(bill_length_mm, na.rm=T)) %>%
  ggplot()+
  geom_col(aes(x=species, y=avg),
           fill="blue")
```

> See [here](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf) for a list of R color names.

You can also use other color palettes by installing `viridis` or `hrbrmst themes`


### Labels

You can add labels with `geom_label` or `geom_text`. `geom_text` is just text and `geom_label` is text inside a rounded white box (this, of course, can be changed). Compare:

```{r}
ggplot(rsurvey)+       
  geom_bar(aes(x= qr_experience, fill=qr_experience))+
  geom_text(aes(x= qr_experience,
                label=..count..),
            stat='count', color="black")+
  labs(title="Using geom_text")

ggplot(rsurvey)+       
  geom_bar(aes(x= qr_experience, fill=qr_experience))+
  geom_label(aes(x= qr_experience,
                label=..count..),
            stat='count')+
  labs(title="Using geom_label")
```

*Note*: Because there is no `y` value, these graphics use `y=..count..` to get the total number and `stat="count"` to say you will use the sum in the aesthetic.

### Multiple Plots

#### Faceting

You can break a graphic into multiple plots (or facets) using `facet_wrap(~variable)`. Here is an example:

```{r message=FALSE, warning=FALSE}
ggplot(airpassengers)+
  geom_line(aes(x=year, y=AirPassengers, group=month))+
  facet_wrap(~month)
```

*Note:* The months are not in order. To put them in order, you would first need to use `factor()` inside a `mutate()` command.

You can control the **number of rows/columns** use `nrow=` or `ncol=`:

```{r message=FALSE, warning=FALSE}
airpassengers %>%
  mutate(month = factor(month, levels=c(
    "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Oct",
    "Nov", "Dec"
  ))) %>%
ggplot()+
  geom_line(aes(x=year, y=AirPassengers, group=month))+
  facet_wrap(~month, ncol=3)+
  labs(title="Months ordered and placed in 3 columns")
```


#### `patchwork`

You can also use the `patchwork` package to connect different plots using `+`, `/`, and `|`.


First, save your plots as data objects:

```{r message=FALSE, warning=FALSE}
library(patchwork)

#example plots
p1 <- ggplot(mtcars) + 
  geom_point(aes(mpg, disp)) + 
  ggtitle('Plot 1')

p2 <- ggplot(mtcars) + 
  geom_boxplot(aes(gear, disp, group = gear)) + 
  ggtitle('Plot 2')

p3 <- ggplot(mtcars) + 
  geom_point(aes(hp, wt, colour = mpg)) + 
  ggtitle('Plot 3')

p4 <- ggplot(mtcars) + 
  geom_bar(aes(gear)) + 
  facet_wrap(~cyl) + 
  ggtitle('Plot 4')
```

##### Using `+` - Side-by-Side

```{r}
p1 + p2
```

##### Using `/` - Stacked

```{r}
p1 / p2
```

##### Using `|` - Nested

```{r}
p1 | (p2 / p3)
```

#### Controlling layout

```{r}
p1 + p2 + p3 + p4 + 
  plot_layout(widths = c(2, 1))
```

> See [https://github.com/thomasp85/patchwork](https://github.com/thomasp85/patchwork) for more information.

### Themes

Themes control the overall look and feel of `ggplot`. If there is a specific theme, it is called using `theme_name()`. If you are modifying theme elements, you will use `theme()`

#### Pre-Installed Themes

Here are some examples:

```{r}
ggplot(mtcars) + 
  geom_point(aes(mpg, disp)) + 
  ggtitle('No theme specified')

ggplot(mtcars) + 
  geom_point(aes(mpg, disp)) + 
  theme_void()+
  ggtitle('theme_void()')

ggplot(mtcars) + 
  geom_point(aes(mpg, disp)) + 
  theme_classic()+
  ggtitle('theme_classic()')

ggplot(mtcars) + 
  geom_point(aes(mpg, disp)) + 
  theme_minimal()+
  ggtitle('theme_minimal()')
```

#### `ggthemes` and other theme packages

`ggthemes` and `hrbrthemes` are two popular theme packages. Here are some examples:

##### `ggthemes`

```{r}
library(ggthemes)


ggplot(mtcars) + 
  geom_point(aes(mpg, disp)) +
  theme_wsj()+
  ggtitle('theme_wsj()')

ggplot(mtcars) + 
  geom_point(aes(mpg, disp)) +
  theme_tufte()+
  ggtitle('theme_tufte()')
```

> See all themes and more at [https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/)

##### `hrbrthemes`

```{r message=FALSE, warning=FALSE}
library(hrbrthemes)

ggplot(mtcars) + 
  geom_point(aes(mpg, disp)) +
  theme_modern_rc()+
  ggtitle('theme_modern_rc()')
```

> See more themes are color scales here: [https://github.com/hrbrmstr/hrbrthemes](https://github.com/hrbrmstr/hrbrthemes)

#### Controlling theme elements

If you simply want to remove lines, change the legend position, etc, you can use `theme()`. Here are two quick examples.

```{r}
ggplot(mtcars) + 
  geom_point(aes(mpg, disp, color=as.factor(gear)))+
  labs(title="No theming",
       color="Gears")

ggplot(mtcars) + 
  geom_point(aes(mpg, disp, color=as.factor(gear)))+
  labs(title="Several theme changes",
       color="Gears")+
  theme(
    panel.background = element_blank(),
    legend.position = "top"
  )
```

> There are a lot of ways you can customize a theme. See [https://ggplot2.tidyverse.org/reference/theme.html](https://ggplot2.tidyverse.org/reference/theme.html).